# Reproducible Research: Peer Assessment 1
This assignment makes use of data from a personal activity monitoring device.
This device collects data at 5 minute intervals through out the day. The data
consists of two months of data from an anonymous individual collected during
the months of October and November, 2012 and include the number of steps
taken in 5 minute intervals each day.

## Loading and preprocessing the data
The data file (activity.zip) was unzipped into the default directory.  The data file (activity.csv) was loaded with read.csv with no parameters.  

summary() revieled 2304 NAs.  

str() shows the dataset includes the following columns:  steps (int), date (Factor) and interval(int). 

For immediate analysis, NAs were removed, creating activeComplete. A day of the week was derived from the date using weekdays() to facilitate additional subsetting and analysis.


```{r}
setwd("C:\\Users\\srobin\\Documents\\GitHub\\RepData_PeerAssessment1")
unzip(paste0(getwd(),"\\","activity.zip"))
active <- read.csv(paste0(getwd(),"\\","activity.csv"))

summary(active)

str(active)

activeComplete <- active[complete.cases(active),]
activeComplete$dow <- weekdays(as.Date(activeComplete$date))
summary(activeComplete)

```


## What is mean total number of steps taken per day?

Complete cases was aggregated by date to derive 53 days of observations and summed for each day.  A histogram shows the frequency of the steps per day was around 10,000, closely matching a summary of this data, with a mean of 10,766 and a median of 10,765.

```{r echo=FALSE}
## histogram of the total number of steps taken each day
perDay <- activeComplete[,1:2]
sum_steps_per_Day <- aggregate(.~ date, data=perDay,sum)

hist(sum_steps_per_Day$steps, plot=TRUE, main="Histogram of Total Steps each Day")
# summary(sum_steps_per_Day)

```



```{r echo=FALSE}
## mean and median number of steps taken each day

mSteps_per_Day <- aggregate(.~ date, data=perDay,sum)
mSteps <- mean(mSteps_per_Day$steps)
medSteps <- median(mSteps_per_Day$steps)

```

### Mean and Median  
- Mean: `r mSteps`
- Median:  `r medSteps`

## What is the average daily activity pattern?

The average daily activity shows increasing activity after Interval 500 (5:00am) and a sharp spike around Interval 800 (8:00am).  Maximum activity occurs at Interval 835 (8:30am).  Activity fluxuates across the day, until around Interval 1900 (7:00pm), where it declines quickly and tappers off after 2000 (10:00pm).

```{r}
## time series plot of the average number of steps taken 
## (averaged across all days) versus the 5-minute intervals

daily <- activeComplete[,c(1,3)]
  pattern <- aggregate(.~ interval,data=daily, mean, na.rm=TRUE)
  plot(pattern$interval,pattern$steps, type="l", main="Time Series of Average \n Number of Steps per Day")

```

###Average Maximum number of steps
```{r echo=FALSE}
## 5-minute interval that, on average, contains the maximum number of steps

  maxSteps <- subset(pattern, steps == max(pattern$steps))
  maxSteps

```



## Inputing missing values

Where data was missing steps (NA), steps were estimated by using the mean of the corresponding interval and day of the week of records that had complete data.  A dataset of complete cases was created, along with Weekdays() function to create a 'Day-of-the-Week'column (dow). 

The dataset was grouped by interval + dow, creating a mean of each interval by day. Missing data is replaced day by day through a loop of missing dates.

```{r echo=FALSE}

## strategy for imputing missing data
# Just for fun
# library(mice)
# miced <- mice(active)
# library(Amelia)
# a <- amelia(x=active,idvars="date")
# head(a$imputations$imp5)

# Duplicate dataframe to be used to append replacement data
replaceMissing <- active[complete.cases(active),]
replaceMissing$dow <- weekdays(as.Date(replaceMissing$date))

# Replacement data
mean_steps_per_Day <- aggregate(replaceMissing$steps, 
                                by=list(interval = replaceMissing$interval,
                                        dow = replaceMissing$dow),
                                            FUN=mean)


dowMean <- mean_steps_per_Day[,1:3]
names(dowMean)[1:3] <-  c("interval","dow","steps")

# Dataset where steps are NA
missing <- subset(active, is.na(active$steps))
missing$dow <- weekdays(as.Date(missing$date))

# get the Dates and DOW where data is missing
miss <- aggregate(missing$steps,by=list(date=missing$date),mean)
miss$dow <- weekdays(as.Date(miss$date))



for(i in 1:nrow(miss)){

  mUse <- subset(missing,missing$date == miss$date[i])
  iData <- subset(mean_steps_per_Day, mean_steps_per_Day$dow == miss$dow[i])
  names(iData)[3] <- "steps"
  
  mUse$steps <- ifelse(is.na(mUse$steps),
                          iData$steps[mUse$interval==iData$interval] ,
                          mUse$steps)
  
  # merge missing and activeComplete 
  replaceMissing <- rbind(mUse,replaceMissing)
 
  
}


meanActiveComplete <- mean(activeComplete$steps)
meanReplaceMissing <- mean(replaceMissing$steps)



```
```{r}
summary(replaceMissing)
```
The introduction of Replaced data has slightly raised the mean,   
- from `r meanActiveComplete`  
- to `r meanReplaceMissing`

## Are there differences in activity patterns between weekdays and weekends?
Subject seems to "sleep-in"" during the weekends, activity doesn't ramp up until around 7:00am, but has a more active pattern during the rest of the day.


```{r echo=FALSE}

## histogram of the total number of steps taken each day after missing values were 
## imputed
# create a factor for Weekend - Weekend or Weekday
replaceMissing$weekend <- factor(ifelse(replaceMissing$dow %in% c("Saturday", "Sunday"), "Weekend", "Weekday"), ordered = FALSE)

meanWeek <- aggregate(replaceMissing$steps, 
                                 by=list(interval=replaceMissing$interval, 
                                         weekend=replaceMissing$weekend),  mean)

#  Reorder for facet
meanWeek$weekend2 <- factor(meanWeek$weekend, levels=c("Weekend", "Weekday"))


library(ggplot2)

ggplot(meanWeek, aes(x=interval, y=x)) + 
  geom_line(colour="#0000FF") + 
    facet_wrap(~weekend2, ncol=1) +
  theme(panel.background = element_rect(fill = 'white')) +
  theme(strip.text.x = element_text(size=8),
        strip.text.y = element_text(size=12, face="bold"),
        strip.background = element_rect(colour="white", fill="#FaEFEF")) +
  ylab("Number of Steps") 



```



